Statistical Computing (MATH6173) Coursework 2

### 2023-01-04 draft_report.Rmd
3rd review of report, slightly cleaning it up (one/two more small reviews needed)

For acceptance-rejection sampling (A-R sampling) we need to find a constant $M>0$ such that $\pi(x) \leq Mg(x), \forall x \in \mathbb{R}$. For $\pi(x)$ the target density (the density from which we want to generate a sample of random values) and $g(x)$ the proposal density defined in equations 1 and 2 above respectively. This constant M determines the acceptance probability which is $\frac{1}{M}$ and hence the efficiency of the algorithm. For multiple possible values of $\lambda$ in $g(x)$ the value of $\lambda$ that results in a smaller constant M will have a higher acceptance probability and thus will be our preferred value of $\lambda$ for the algorithm. To find M for each of the proposed $\lambda$ values I split the real domain of $x$ into 3 regions (see plot 1) that I call 1,2 and 3 which correspond to the different expressions of $\pi(x)$ depending on the value of $x$. Then study the relationship between $g(x)$ and $\pi(x)$ within these 3 regions to find $M_{1},M_{2},M_{3}$ which correspond to the smallest constant such that $\pi(x) \leq M_{i}g(x)$ for all possibles values of x in the ith region. Then set $M=max(M_{1},M_{2},M_{3})$ as this is the smallest constant that satisfies $\pi(x) \leq Mg(x)$ for all possible values of $x \in \mathbb{R}$.

#### Region 1:
For the first region ($-\infty, s-b$), $\pi(x)=0$ thus we can let $M_{1}$ be some arbitrarily small value greater than 0 say $\epsilon$. This will satisfy $\pi(x) \leq M_{1}g(x)$ in the region specified as $g(x) > 0$ for all values of $x$ in this region.

#### Region 2:
For the second region [$s-b,s$], $\pi(x)=\frac{1}{2b}$. Considering that $g(x)$ is an increasing function in this region (see plot 1) it will be at it's minimum value at $x=s-b$ and the value of $\pi(x)$ is constant. Thus, setting $M_{2}=\frac{\pi(s-b)}{g(s-b)}=\frac{exp(\lambda|s-2b|)}{\lambda b}$ will satisfy the equation $\pi(x) \leq M_{2}g(x)$ for all values of  $x$ in this region.

#### Region 3:
For the third region ($s,\infty$), $\pi(x)=\frac{as^{a}}{2x^{a+1}}$. I note that $\pi(x)$ is a decreasing function in this region (see plot 1) and initially $g(x)$ is a smaller but increasing function (see plot 1), until it reaches its maximum value at which point it is larger than $\pi(x)$. It then becomes a decreasing function of $x$ but is always larger than $\pi(x)$. Thus, setting $M_{3}=\frac{\pi(s)}{g(s)}=\frac{a}{\lambda s}exp(\lambda |s-b|)$ will satisfy the equation $\pi(x) \leq M_{3}g(x)$ for all $x$ in this region.

@@

#### Sampling from the proposal distribution:
To sample from $g(x)$, use the inverse transformation technique by drawing $u$ from a Uniform(0,1) distribution then if $u<0.5$ set $x=b+\frac{log(2u)}{\lambda}$, else set $x=b-\frac{2-2u}{\lambda}$. To obtain the value $x$, drawn from $g(x)$ the proposal distribution.

@@

* $s-b <x$, check if $y \leq \frac{\pi(x)}{Mg(x)}=\frac{1}{2b}\frac{2}{M\lambda exp(-\lambda|x-b|)}$

If either of these checks are true then $x$ is a draw from the target distribution $\pi(x)$ and the value is saved, if these checks are not true or $x<s-b$ then $x$ is not a draw from $\pi(x)$ and the value of $x$ is disregarded. Repeat this process until the desired number of samples from $\pi(x)$ has been attained.

@@

\begin{equation*}
Pr(U>\boldsymbol{b})=\prod_{i=1}^{d}Pr\left(Z>\frac{b_{i}-u_{i-1}}{\sqrt{t_{i}-t_{i-1}}}\right)=\prod_{i=1}^{d}\int_{s_{i}}^{\infty}\phi(x)dx=\prod_{i=1}^{d}\int_{-\infty}^{\infty}\frac{\phi(x)}{q_{i}(x)}q_{i}(x)I(x>s_{i})dx = E\left[\prod_{i=1}^{d} \frac{\phi(V_{i})}{q_{i}(V_{i})}I(V_{i}>s_{i})\right]
\end{equation*}

With $V_{i}, i \in \{1,..,d\}$ i.i.d. random variables from the proposal distribution, $q_{i}$ the density of the proposal distribution, $\phi$ the density of a standard normal distribution and $I$ the indicator function.

@@

Where $w_{i}^{(j)}$ is $\frac{\phi(V_{i})}{q_{i}(V_{i})}$ on the jth iteration and the superscript $(j)$ denoting the jth sampling. Therefore, to not waste any sample values let the proposal distributions be exponential distributions shifted by $s_{i}$ as all sample values (denoted $v_{i}$) drawn from these proposal distributions will satisfy $v_{i}>s_{i}$ for $i \in \{1,...,d\}$. Thus, all values generated from these proposal distributions can be used to calculate $I_{n}^{IS}$ from equation (4) as the indicator function is 1 for each $v_{i}$ drawn from each $V_{i}$.

@@

#### Drawing values from proposal distributions:
Next for each iteration j we must simulate the random walk $U$ using equation (3), the initial value $u_{0}$ and times $t_{i}, i \in \{0,..,d\}$ and storing the values of the random walk $\boldsymbol{u}$ as mentioned in part a. Using these values and the relevant proposal distribution of an exponential distribution shifted by $s_{i}$ (described in part a) to generate $v_{i}, i \in \{1,..,d\}$.

@@

Then calculate the product of the weights $\prod_{i=1}^{n}w_{i}$ and save this as the jth value in a vector.

#### Calculating probability:
To calculate $p_{tail}(\boldsymbol{b})$, the previous 2 steps are performed n times thus we have n values of the product of the weights (which I will denote $\hat{w}^{(j)}$ where the value of $j$ denotes the jth iteration) which we then use to calculate the importance sampling (IS) estimator using equation (4) in part a:

@@

#### Drawing new proposal value:
For each iteration j start by drawing $\boldsymbol{P}_{\boldsymbol{T}}^{new}$ the proposed value of $\boldsymbol{P}_{\boldsymbol{T}}$ from the proposal density by drawing each $p_{x}'$ with $x \in \{s,c,r\}$ from a Dirichlet $(20p_{x}^{(cur)})$ distribution and combining these to get $\boldsymbol{P}_{\boldsymbol{T}}^{new}=(p_{s}',p_{c}',p_{r}')^{T}$.

@@

#### Accept or reject proposal:
Calculate $\alpha = min \left( 1,\frac{p(\boldsymbol{P}_{\boldsymbol{T}}^{new}|X,\pi)}{p(\boldsymbol{P}_{\boldsymbol{T}}^{cur}|X,\pi)}\frac{f(\boldsymbol{P}_{\boldsymbol{T}}^{cur})}{f(\boldsymbol{P}_{\boldsymbol{T}}^{new})}\right)$ using the ratios calculated above and generate $u$ a value drawn from a Uniform(0,1) distribution and if $u<=\alpha$ set $\boldsymbol{P}_{\boldsymbol{T}}^{new}$ to be the jth value of $\boldsymbol{P}_{\boldsymbol{T}}$ and update $\boldsymbol{P}_{\boldsymbol{T}}^{cur}$ to be the value of $\boldsymbol{P}_{\boldsymbol{T}}^{new}$. Otherwise, set $\boldsymbol{P}_{\boldsymbol{T}}^{cur}$ to be jth value of $\boldsymbol{P}_{\boldsymbol{T}}$ and don't update the value of $\boldsymbol{P}_{\boldsymbol{T}}^{cur}$.

#### Iterate:
Repeat until we have attained n+b+1 sample values of $\boldsymbol{P}_{\boldsymbol{T}}$ where n is the number of desired samples and b is the specified burn in number. Then throw away the first b+1 values of $\boldsymbol{P}_{\boldsymbol{T}}$. The remaining n values of $\boldsymbol{P}_{\boldsymbol{T}}$ is the estimated posterior distribution of $\boldsymbol{P}_{\boldsymbol{T}}$.

>>>>

For acceptance-rejection sampling (A-R sampling) we need to find a constant $M>0$ such that $\pi(x) \leq Mg(x), \forall x \in \mathbb{R}$. With $\pi(x)$ the target density (the density from which we want to generate a sample of random values) and $g(x)$ the proposal density defined in equations 1 and 2 above respectively. This constant M determines the acceptance probability which is $\frac{1}{M}$ and hence the efficiency of the algorithm. For multiple possible values of $\lambda$ in $g(x)$ the value of $\lambda$ that results in a smaller constant M will have a higher acceptance probability and thus will be our preferred value of $\lambda$ for the algorithm. To find M for each of the proposed $\lambda$ values I split the real domain of $x$ into 3 regions (see plot 1) that I call 1,2 and 3 which correspond to the different expressions of $\pi(x)$ depending on the value of $x$. Then study the relationship between $g(x)$ and $\pi(x)$ within these 3 regions to find $M_{1},M_{2},M_{3}$ which correspond to the smallest constant such that $\pi(x) \leq M_{i}g(x)$ for all possibles values of $x$ in the ith region. Then set $M=max(M_{1},M_{2},M_{3})$ as this is the smallest constant that satisfies $\pi(x) \leq Mg(x)$ for all possible values of $x \in \mathbb{R}$.

#### Region 1:
For the first region $x \in (-\infty, s-b)$, $\pi(x)=0$ thus we can let $M_{1}$ be some arbitrarily small value greater than 0 say $\epsilon$. This will satisfy $\pi(x) \leq M_{1}g(x)$ in the region specified as $g(x) > 0$ for all values of $x$ in this region.

#### Region 2:
For the second region $x \in [s-b,s]$, $\pi(x)=\frac{1}{2b}$. Considering that $g(x)$ is an increasing function in this region (see plot 1) it will be at it's minimum value at $x=s-b$ and the value of $\pi(x)$ is constant. Thus, setting $M_{2}=\frac{\pi(s-b)}{g(s-b)}=\frac{exp(\lambda|s-2b|)}{\lambda b}$ will satisfy the equation $\pi(x) \leq M_{2}g(x)$ for all values of  $x$ in this region.

#### Region 3:
For the third region $x \in (s,\infty)$, $\pi(x)=\frac{as^{a}}{2x^{a+1}}$. I note that $\pi(x)$ is a decreasing function in this region (see plot 1) and initially $g(x)$ is a smaller but increasing function (see plot 1), until it reaches its maximum value at which point it is larger than $\pi(x)$. It then becomes a decreasing function of $x$ but is always larger than $\pi(x)$. Thus, setting $M_{3}=\frac{\pi(s)}{g(s)}=\frac{a}{\lambda s}exp(\lambda |s-b|)$ will satisfy the equation $\pi(x) \leq M_{3}g(x)$ for all $x$ in this region.

@@

#### Sampling from the proposal distribution:
To sample from the proposal distribution, use the inverse transformation technique by drawing $u$ from a Uniform(0,1) distribution then if $u<0.5$ set $x=b+\frac{log(2u)}{\lambda}$, else set $x=b-\frac{2-2u}{\lambda}$. To obtain the value $x$, drawn from the proposal distribution.

@@

* $s-b \leq x \leq s$, check if $y \leq \frac{\pi(x)}{Mg(x)}=\frac{1}{2b}\frac{2}{M\lambda exp(-\lambda|x-b|)}$

If either of these checks are true then $x$ is a draw from the target distribution with density $\pi(x)$ and the value is saved, if these checks are not true or $x<s-b$ then $x$ is not a draw from $\pi(x)$ and the value of $x$ is disregarded. Repeat this process until the desired number of samples from the target distribution has been attained.

@@

\begin{equation*}
Pr(U>\boldsymbol{b})=\prod_{i=1}^{d}Pr\left(Z>\frac{b_{i}-u_{i-1}}{\sqrt{t_{i}-t_{i-1}}}\right)=\prod_{i=1}^{d}\int_{s_{i}}^{\infty}\phi(x)dx=\prod_{i=1}^{d}\int_{-\infty}^{\infty}\frac{\phi(x)}{q_{i}(x)}q_{i}(x)I(x>s_{i})dx = E\left[\prod_{i=1}^{d} w_{i}(V_{i})I(V_{i}>s_{i})\right]
\end{equation*}

With $V_{i}, i \in \{1,..,d\}$ i.i.d. random variables from the proposal distribution, $q_{i}$ the density of the proposal distribution, $\phi$ the density of a standard normal distribution, $w_{i}(V_{i})$ the (unnormalized) importance weights (equal to $\frac{\phi(V_{i})}{q_{i}(V_{i})}$) and $I$ the indicator function.

@@

Where $w_{i}^{(j)}$ is $\frac{\phi(V_{i})}{q_{i}(V_{i})}$ the superscript $(j)$ denotes the jth sampling iteration. Therefore, to not waste any sample values let the proposal distributions be exponential distributions shifted by $s_{i}$ as all sample values (denoted $v_{i}$) drawn from these proposal distributions will satisfy $v_{i}>s_{i}$ for $i \in \{1,...,d\}$. Thus, all values generated from these proposal distributions can be used to calculate $I_{n}^{IS}$ from equation (4) as the indicator function is 1 for each $v_{i}$ drawn from each $V_{i}$.

@@

#### Drawing values from proposal distributions:
Next for each iteration j we must simulate the random walk $U$ using equation (3), the initial value $u_{0}$ and times $t_{i}, i \in \{0,..,d\}$ and storing the values of the generated random walk $\boldsymbol{u}$ as mentioned in part a. Using these values and the relevant proposal distribution of an exponential distribution shifted by $s_{i}$ (described in part a) then generate $v_{i}, i \in \{1,..,d\}$.

@@

Then calculate the product of the weights for this jth iteration $\hat{w}^{(j)}:= \prod_{i=1}^{n}w_{i}^{(j)}$ and save this value in a vector in element j.

#### Calculating probability:
To calculate $p_{tail}(\boldsymbol{a})$, the previous 2 steps are performed n times thus we have n values of the product of the weights which we then use to calculate the importance sampling (IS) estimator using equation (4) in part a:

@@

#### Drawing new proposal value:
For each iteration j start by generating $\boldsymbol{P}_{\boldsymbol{T}}^{new}$ the proposed value of $\boldsymbol{P}_{\boldsymbol{T}}$ from the proposal density by drawing each $p_{x}'$ with $x \in \{s,c,r\}$ from a Dirichlet $(20p_{x}^{(cur)})$ distribution and combining these to get $\boldsymbol{P}_{\boldsymbol{T}}^{new}=(p_{s}',p_{c}',p_{r}')^{T}$.

@@

#### Accept or reject proposal:
Calculate $\alpha = min \left( 1,\frac{p(\boldsymbol{P}_{\boldsymbol{T}}^{new}|X,\pi)}{p(\boldsymbol{P}_{\boldsymbol{T}}^{cur}|X,\pi)}\frac{f(\boldsymbol{P}_{\boldsymbol{T}}^{cur})}{f(\boldsymbol{P}_{\boldsymbol{T}}^{new})}\right)$ using the ratios calculated above and generate $u$ a value drawn from a Uniform(0,1) distribution and if $u<=\alpha$ set $\boldsymbol{P}_{\boldsymbol{T}}^{new}$ to be the jth sample value of $\boldsymbol{P}_{\boldsymbol{T}}$ and update $\boldsymbol{P}_{\boldsymbol{T}}^{cur}$ to be the value of $\boldsymbol{P}_{\boldsymbol{T}}^{new}$. Otherwise, set $\boldsymbol{P}_{\boldsymbol{T}}^{cur}$ to be jth sample value of $\boldsymbol{P}_{\boldsymbol{T}}$ and don't update the value of $\boldsymbol{P}_{\boldsymbol{T}}^{cur}$.

#### Iterate:
Repeat until we have attained n+b+1 sample values of $\boldsymbol{P}_{\boldsymbol{T}}$ where n is the number of desired posterior sample values and b is the specified burn in number. Then throw away the first b+1 values of $\boldsymbol{P}_{\boldsymbol{T}}$. The remaining n values of $\boldsymbol{P}_{\boldsymbol{T}}$ is the estimated posterior distribution of $\boldsymbol{P}_{\boldsymbol{T}}$.
